Difference Scores: Paired Samples T-test

### 2024-05-01 examples/1a_Examples_DifferenceScores_PairedTtests.qmd
update toolbar
example 1a update

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
theme: cosmo
---

## Overview

<p>

This example assesses growth in a subsample of ABCD participants from Baseline (T0) to the 1-Year follow-up (T1), using weight as a representative metric. The analysis is conducted in two primary steps: 1) a difference score is calculated between baseline and Year_1 weight measurements for each participant; 2) a one-sample t-test is used to test whether the difference score is statistically different than a null hypothesis of zero change. Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's weight at Baseline and Year_1. The ensuing analysis and interpretations are detailed in the subsequent sections.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse", "easystats", "lme4", "gtsummary", "report", "broom", "gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse) # Collection of R packages for data science
library(easystats) # Collection of R packages statistical modeling, visualization, and reporting
library(gtsummary) # Publication-ready tables
library(report) # Easy reporting of regression analyses
library(broom) # Tidy and augment statistical models output
library(gridExtra) # Arrange multiple grid-based plots on a page

```
:::

```{r config}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

### Config Options
#::: blue
#**This code configures knitr code chunk options**

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue

**This code reads in and shows the data to be used in the current example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroweightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
    full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c(
    "baseline_year_1_arm_1",
    "1_year_follow_up_y_arm_1",
    "2_year_follow_up_y_arm_1",
    "3_year_follow_up_y_arm_1",
    "4_year_follow_up_y_arm_1"
)

df <- merged_data %>%
    filter(eventname %in% eventnames_to_include) %>%
    mutate(
        src_subject_id = as.factor(src_subject_id),
        eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
        age = as.numeric(age),
        sex = as.factor(sex),
        race.4level = as.factor(race.4level),
        hisp = as.factor(hisp),
        high.educ.bl = as.factor(high.educ.bl),
        household.income.bl = as.factor(household.income.bl),
        acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
        rel_family_id.bl = as.factor(rel_family_id.bl),
        site_id_l = as.factor(site_id_l),
        nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
        anthroweightcalc = as.numeric(anthroweightcalc),
        anthroweightcalc = as.numeric(anthroweightcalc)
    ) %>%
    # Exclude cases from unused assessment waves
    filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
    select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
    mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1", "Year 2", "Year 3", "Year 4"))) %>%
    mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
    tbl_summary(
        by = eventname,
        missing = "no",
        label = list(
            sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic",
            anthroweightcalc ~ "Weight"
        ),
        statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```

:::

### Results

::: blue
The code snippet below tells R to compute a difference score by subtracting each participant's weight at T1 from their weight at T0. Relevant summary statistics are also provided.

**STEP 1: Compute Difference Score**

```{r build-model-ttest-1}
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| class.source: 'fold-hide'

# Function to compute difference scores
compute_difference <- function(df, variable_name) {
    # Define the event names of interest
    baseline_event <- "baseline_year_1_arm_1"
    followup_event <- "1_year_follow_up_y_arm_1"

    # Compute the difference between Baseline and Year 1 data for the given variable
    diff_data <- df %>%
        filter(eventname %in% c(baseline_event, followup_event)) %>% # Filter for specific event names
        select(src_subject_id, eventname, all_of(variable_name)) %>% # Select required columns
        spread(eventname, variable_name) %>% # Convert data from long to wide format
        mutate(diff = get(followup_event) - get(baseline_event)) %>% # Compute difference between the two time points
        mutate(group_resid = (diff) - mean(diff)) %>% # compute residuals (group_resid).
        drop_na(diff) # Exclude rows with NA in the computed difference

    # Summarize the computed difference scores
    diff_summary <- summary(diff_data$diff)

    # Return the difference data and its summary
    list(data = diff_data, summary = diff_summary)
}

# List of variables for which difference scores are to be computed
variables_of_interest <- c("anthroweightcalc")

# Compute and store difference scores and summaries for each variable in a list
difference_and_summary_list <- lapply(variables_of_interest, function(var) {
    compute_difference(df, var)
})

# Extract the difference data for the 'anthroweightcalc' variable
weight_diff_data <- difference_and_summary_list[[1]]$data

# Merge the 'diff' column back to the main df using 'src_subject_id' as the key
df <- left_join(df, weight_diff_data %>% select(src_subject_id, diff), by = "src_subject_id")

@@

:::
:::

>>>>

---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, ehi_y_ss_scoreb, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

@@

## Overview
Difference scores can be used to assess change in a variable across two timepoints. The value of the variable at the first timepoint is substracted from the value at the second timepoint. A one-sample t-test can then be applied to determine whether the mean change is significantly different from zero.

This example will examine height values obtained across 2 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to determine whether height values in the sample, on average, increase across the two timempoints. A difference score will be calculated between participant's height values collected at baseline and again at the 1-year follow-up. Next, a one-sample t-test will be used to test whether the difference score is statistically different than a null hypothesis of zero change. Finally, a visual inspection is further conducted via a scatterplot to graphically represent the relationship between participant's height across assessments.

## Preliminary Setup
### Load R libraries
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

## Results
### Compute Difference Score
This code computes a difference score by subtracting participant's height at baseline from height at the 1-year follow-up. Next, a one-sample t-test is conducted to determine whether the mean difference score is different from zero. Model output is provided below, as well as a brief summary of results.

```{r}

# Reshaping data from long to wide
df_wide <- df_long %>%
    pivot_wider(
        id_cols = id,
        names_from = event,
        values_from = height,
        names_prefix = "Height_"
    )

# Compute the height difference score
df_wide <- df_wide %>%
    mutate(
        height_diff = Height_Year_1 - Height_Baseline
    )

# Calculate summary statistics for the difference score
summary(df_wide$height_diff)

# Compute summary statistics for Height by eventname
summary <- df_long %>%
    group_by(event) %>%
    get_summary_stats(height, type = "mean_sd")

data.frame(summary)

The summary statistics provide insights into participant's height across two annual assessments. At baseline, the average height is 55.24 inches (sd = 3.33). Over the year, there is an increase of 2.35 inches in average height to 57.59 inches at the time of the 1-year follow-up.

### Conduct t-test on Difference Score
A one-sample t-test is computed on the difference scores to examine whether there is evidence of change in participant's height between the two timepoints.
```{r}

# Perform a one-sample t-test on the difference scores for height
model <- t.test(df_wide$height_diff, mu = 0)
model

The results from the one-sample t-test conducted on the height difference scores reveals an increase in height from Baseline to Year 1. The t-test yields a t-value of 98.86 with a p-value less than <.001, indicating that the average height increase of 2.36 inches is significantly different from zero. Additionally, the 95% confidence interval indicates that the true mean difference score in the population lies between approximately 2.31 to 2.41 inches.

### Model Plots
```{r}

# Scatterplot to compare height differences across two events
ggplot(df_wide, aes(x = Height_Baseline, y = Height_Year_1)) +
    geom_point(aes(color = Height_Baseline), alpha = 0.6) + # Color points by event type, adjust for your data
    labs(
        x = "Height at Baseline",
        y = "Height at Year 1",
        title = "Scatterplot of Heights at Baseline and Year 1",
        subtitle = "Each point represents a participant"
    ) +
    theme_minimal() +
    geom_smooth(method = "lm", se = FALSE) + # Add a regression line without confidence interval
    theme(legend.position = "bottom")

The scatterplot visually depicts the relationship between individuals' heights at baseline and their heights at year 1. Each point on the plot represents an individual, with their baseline height plotted on the x-axis and their year 1 height on the y-axis. A noticeable positive linear trend can be observed, as highlighted by the blue regression line, indicating that those who were taller at baseline generally remained taller at year 1. The strength and direction of this relationship suggests a strong positive association between baseline and year 1 height values.

## Wrapping Up
This analysis examined difference scores of participant's height using a one-sample t-test. Findings showed a significant increase in height values over the one-year interval of 2.36 inches, with confidence intervals ranging from 2.31 to 2.41 (t = 98.864, df = 11135, p-value < 2.2e-16). Further, a scatterplot visualizing the relationship between baseline and Year_1 weights showed a strong positive linear trend. This suggests that participants who were taller at baseline generally remained taller at Year_1, reaffirming the consistent growth trend observed in the data.

This difference score approach used in this example quantifies the change in height values over time within a single group and uses a one-sample t-test to test whether this value differs significanctly from 0. More generally, this approach is often implemented when data used to evalaute change is only available at two timepoints, as more flexible approaches are typically implemented when data from additional (>2) time points is available.
